Substitution rate responses to changes in effective population size.

### 2020-04-22 manuscript/manuscript.tex
Nicolas corrections

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1
\captionsetup{width=0.85\textwidth}
\include{notations}
\begin{document}
\title{Substitution rate responses to changes in effective population size.}
\author{T. Latrille, N. Lartillot}
\maketitle

\abstract{
The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
Such theoretical prediction had been observed in empirical data across many clades.
However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
We apply our framework in the special case of fitness proportional to the probability of protein folding.
Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.

@@

Molecular sequences differ across species due to the particular history of DNA substitutions along each specie's respective lineage.
These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
These forces have effects at different levels, mutations are carried by molecular sequence, selection is beared by individual's fitness, and random genetic drift is a population effect.
The strength of random genetic genetic, or stochastic sampling of mutations, is less pronounced in lineage with high effective population size ($\Ne$), and as a consequence the purification by selection of deleterious mutations is more effective.
Such idea has emerged in the framework of quasi-neutral theory of evolution, whereby predicting a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
This prediction is valid under the assumption that most mutations are deleterious and that their selective effect on individuals are drawn from a fixed distribution \cite{Welch2008}.
Furthermore, this prediction assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}.
This relative substitution rate of selected mutations, called $\dnds$ is this study, is a theoretical macroscopic observable in molecular sequence evolution, given we can separate selected and neutral mutations.
Practically, $\dnds$ can be approximated in protein-coding DNA sequences, where selected mutations are the non-synonymous changes affecting the amino-acid sequence, and the non-synonymous mutations are supposedly neutral, this statistic is called $dNdS$.\\
% \dnds decrease with \Ne under the assumption of fixed DFE

The context of protein coding sequence fostered another modeling of selection, based on genotype-fitness map instead of distribution of fitness effects.
In such approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
For example, if the target amino-acid has an higher fitness, the selective effect of the mutation is positive, and reciprocally negative for lower fitness.
Such modeling approach account for higher substitutions rate between amino-acid with similar bio-chemical properties.
Fundamentally, in such framework the fitness depends solely on the actual genotype, not on the trajectory of mutations leading to this sequence.
Even thought the modeling approach is substantially different, assuming each site of the sequence has its own genotype-fitness map also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.\\
% \dnds decrease with \Ne under the assumption site fitness landscape

Empirically, fluctuation of $\dnds$ along lineages of phylogenies have been inferred using protein-coding DNA alignments \cite{Yang2001, Zhang2004}.
Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have asserted that $\dnds$ correlates positively with longevity and body mass, traits which have been sampled only in the extant lineages \cite{Lartillot2011,Weber2014}.
Moreover, longevity and body mass are proxies of $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}.
As such, empirical evidences suggest a negative correlation between $\dnds$ and $\Ne$, confirming the theoretical prediction of the quasi-neutral theory of evolution.
However, this correlation could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation could be in the inverse direction or not statistically significant \cite{Figuet2016}.\\
% \dnds decrease with \Ne empirically, but not always

Mitigated empirical evidence that $\dnds$ decrease with $\Ne$ encouraged alternative explanatory mechanism \cite{Lanfear2014}.
One striking result proved that $\dnds$ can independent of $\Ne$, when fitness is a function of a phenotype and that such phenotype is equimutable.
Equimutability states that the probabilities of phenotype changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
Such argument has been invoked in the \textit{in silico} observation that $\dnds$ is seamingly independent of $\Ne$ if the protein thermodynamic stability is computed using a 3D structural model of the protein, and that the fitness is proportional to the folding probability of the protein \cite{Goldstein2013}.
The given explanation is that probabilities of changes in free energy of folding ($\deltadeltaG$) is independent of the current free energy ($\deltaG$), which is considered an equimutable phenotype.\\
% Explanation could be that \dnds is independent of \Ne

However, this assumption is relatively strong, as for example if a protein is completely stable, only destabilizing mutations can occur, and more generally the probabilities of destabilizing mutational events is dependent on the actual composition of the protein.
Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
The goal of this study is to derive the $\dnds$ relationship to $\Ne$ to intrinsic molecular parameters of the model.
More generally, we develop mathematical tools to derive the $\dnds$ relationship to $\Ne$ in the context of a genotype-phenotype-fitness map.
With such results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data.\\
% The assumptions are strong, we want to relax equimutability which is unlikely

Orthogonally to the relationship between $\dnds$ and $\Ne$, independent studies have argued that protein abundance is also a predictor of $\dnds$.
As such, we also seek to derive the $\dnds$ relationship to protein abundance, and discuss the empirical implication.\\
% \dnds is relationship to protein abundance

Lastly, the theoretical results discussed so far are all obtained when the balance between mutation, selection and drift is at equilibrium.
However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ leads first to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection \cite{Jones2016}.
Studying only equilibrium properties can thus be misleading, and the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed by our model.\\
% Result at equilibrium, what about dynamical properties?

\section*{Results}
\subsection*{Theoretical approximations}
In this section we present an analytical approximate solution for the change in $\dnds$ at equilibrium after a change in $\Ne$.
The main results are given in a general case of any phenotype-fitness map, and also derived in the specific case of fitness proportional to the total amount of folded protein.
All development are available in Supplementary Materials in the most general case.
The first step define the relationship from genotype to fitness, through an intermediary phenotype.
The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation. \\

We base our work on numerical model of protein stability \cite{Williams2006, Goldstein2011, Pollock2012}, on which we make simplifying assumptions to derive analytical equations.
In the original model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
Technically, the free energy of the folded are unfolded conformations are computed using $3$D conformations of the protein.
In such model, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.\\

We approximate this model such that the destabilizing effect of an amino-acid does not depend on a $3$D representation of the protein, but each site contribute independently to $\deltaG$.
With such approximation, at a given site of sequence each amino-acid is either stabilizing or destabilizing the protein, and an unstable amino-acid contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ compare to a stable amino-acid.
If all amino-acids of the sequence are stable, $\deltaG$ equal to $ \alpha < 0$. 
In this model, the more compact phenotype summarizing the genotype is the proportion of unstable amino-acid in the sequence, defined as $0 \leq \x \leq 1$, and $\deltaG$ is a linear function of $\x$:
\begin{align}
\deltaG (\x) = \alpha + \Nsite \gamma \x,
\end{align}
where $\Nsite$ is the number of sites in the sequence. \\

From a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of unstable amino-acid, or do nothing if the mutation is between both unstable (or equivalently stable) amino-acids.
To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable and only one amino-acid is stable and the $19$ others are unstable.
Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
\begin{gather}
\begin{cases}
\dx &\text{ with probability } 1-x, \\
0 &\text{ with probability } \frac{18 \x }{19}, \\
-\dx &\text{ with probability } \frac{\x}{19}.\\
\end{cases} \label{eq:proba}
\end{gather}
In the extreme case of optimal phenotype ($\x = 0$), only destabilization mutations are proposed.
Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between unstable amino-acids (effect $0$) is proportional to $\x$. 
Once obtained the phenotypic effects of mutations and their probabilities, definition of the phenotype-fitness map is necessary derive the selective effect of mutations.\\

Subsequent equations are valid for the more general case where the Wrightian fitness is a log-concave differentiable function $f$ of phenotype $\x$.
As an example, we develop the argument when the Wrightian fitness of an individual is the proportion of protein in folded conformation.
This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm. 
This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
\begin{align}
f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
\end{align}
where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
\begin{align}
s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(\x) }{\partial \x} \label{eq:s} \\
\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x)} \label{eq:s-unfolded}
\end{align}

With the complete genotype-phenotype-fitness map defined (see Figure \ref{fig:NeChangeInfluence}, left panel), one can study the evolutionary processes unfolds.
Starting from an optimal sequence, destabilizing mutations will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, a point of equilibrium called marginal stability.
Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average.
Formally, and after simplification, the equilibrium phenotype denoted $\x\eq$ is given by the equation:
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
\Rightarrow \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19)  & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}. \label{eq:equilibrium}
\end{align}
This equation cannot be solved explicitly for $\x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $\x\eq$ is given in Figure \ref{fig:NeChangeInfluence} (right panel). \\
\begin{figure*}[htb!]
	\begin{mdframed}
		\centering
		\includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
		\caption{
			\textbf{Theoretical model.}
			Panel A. Illustration of the Genotype-phenotype-fitness map. The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence. The fitness is a decreasing log-concave function of the phenotype.
			Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$. The equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}$. The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate. Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $\x\eq$ (x-axis of the crossing with the blue solid line). On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $\x\eq$. Moreover, changes in $\x\eq$ reflects the changes in $\dnds$ since both are related by a monotonous function (equation \ref{eq:dnds}). The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
		}
		\label{fig:NeChangeInfluence}
	\end{mdframed}
\end{figure*}

Moreover, at equilibrium we can derive the expected substitution rate of mutations changing the sequence, and by normalizying by the substitution rate of neutral mutations, $\dnds$ simply approximates to:
\begin{gather}
\dnds \simeq \x\eq \label{eq:dnds}
\end{gather}
This simple approximation is due to substitutions at unstable sites (proportion $\x\eq$) to one of the other unstable amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).\\

From the equilibrium equation \ref{eq:equilibrium}, differential calculus allow to derive the response of $\x\eq$ and ultimately $\dnds$ after a change in $\Ne$.
We define the elasticity $\chi$ as slope of the $\dnds$ response to changes in $\Ne$ in log scale, which simplify to a compact equation: 
\begin{align}
\chi = \frac{ \der \dnds}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
\Rightarrow \chi & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
\end{align}
The elasticity is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype. To note, this elasticity is strictly negative for decreasing log-concave fitness functions, asserting that $\dnds$ is a decreasing function of $\Ne$.
In the case of fitness equal to the folded proportion, $\chi$ is independent of $\x\eq$, meaning that $\dnds$ is linearly decreasing with $\Ne$ in log space.
Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship, where only $\gamma$ and $\Nsite$ are free parameters. \\

>>>>

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1
\captionsetup{width=0.85\textwidth}
\renewcommand{\baselinestretch}{1.5}

\include{notations}
\begin{document}
	\title{Substitution rate responses to changes in effective population size.}
	\author{T. Latrille, N. Lartillot}
	\maketitle

	\abstract{
		The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
		One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
		In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
		Such theoretical prediction had been observed in empirical data across many clades.
		However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
		Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
		Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
		Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
		Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
		We apply our framework in the special case of fitness proportional to the probability of protein folding.
		Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
		We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
		However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
		We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
	}

@@

	[Concernant l'introduction, j'ai pris la liberté de lisser directement le texte, de rajouter quelques petites phrases ici et là, pour compléter et ajuster le fil de l'argument. Il me semble pour autant que je n'ai pas trop trahi ton fil directeur, du moins dans la première partie. Sur la fin de l'intro, par contre, je pense qu'il faut remanier un peu plus. En particulier, (1) la question de l'articulation avec la question de la relation entre dnds et expression, et (2) le fait de bien terminer toute l'analyse de l'état des lieux avant d'en venir à la contribution spécifique de ce manuscrit; dans la version que tu avais proposé, ces deux aspects avaient tendance à se mélanger un peu sur la fin.]

	[je propose de scinder ton premier paragraphe en deux: un premier de perspective un peu plus générale sur l'enjeu de démêler les forces évolutives à partir des patterns de substitution, un deuxième qui en vient plus spécifiquement au rôle de la dérive génétique, et qui s'articule du coup plus directement avec la question de la théorie quasi-neutre (en passant, en anglais, c'est nearly-neutral, et non quasi-neutral). Du coup, concernant ce premier paragraphe-ci, il faut aller jusqu'au bout de l'argument, et j'ai rajouté deux phrases en ce sens.]
	Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
	These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
	These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
	Yet, they jointly contribute to the long-term molecular evolutionary process.
	Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.

	[The specific role of random drift $\to$ the nearly-neutral argument.]
	One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
	Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
	This fundamental idea is at the core of the nearly-neutral theory of evolution.
	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious [d'abord expliciter cette composante importante de la théorie quasi-neutre].
	As a result, the theory predicts a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}, relative to the neutral substitution rate.
	[In its original version, the nearly-neutral theory also assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}. {\it pas tout à fait vrai: en fait, la théorie invoquait un effet temps de génération, qui était compensé par un effet Ne, résultant en une horloge moléculaire au niveau non-synonyme. Mais cela me parait un peu compliqué. Peut-être zapper ce point?}]

	The relative substitution rate of selected, relative to neutral, mutations (called $\dnds$ is this study) is a theoretical macroscopic observable in molecular sequence evolution [formulation compliquée.. is an observable quantity of the molecular evolutionary process?], at least if we can separately estimate the rate of neutral mutations in a phylogenetic or comparative context. Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\dnds$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$). Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\dnds =  dN/dS$ as a function of long-term changes in $N_e$.

	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects \cite{Welch2008} {\it citer également Kimura, 1979}. Be more specific here: Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $dnds$ as a function of $N_e$ (i.e. $\dnds \sim N_e^{-\beta}$), where $\beta$ is the shape parameter of the DFE. In practice, DFEs are strongly leptokurtic: thus, predicts a weak negative relation between $\dnds$ and $N_e$.

	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
	For example, if the target amino-acid has a higher fitness than the current amino acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
	Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
	More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.

	Empirically, variation in [fluctuation is more in the short term, I would say] $\dnds$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}. [Ici, donner un compte-rendu un peu plus exhaustif des analyses empiriques publiées au sujet de la relation dnds / Ne. Possiblement en deux temps: les analyses non intégratives: Popadin, et d'autres; puis, les approches de type coevol.]
	Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have found that $\dnds$ correlates positively with life-history traits such as longevity and body mass \cite{Lartillot2011,Weber2014}.
	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\dnds$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
	However, this correlation between $\dnds$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (ref?), or even in the opposite direction (i.e. negative between $dnds$ and LHT) \cite{Figuet2016} (other refs?).

	Mitigated empirical evidence about the response of $\dnds$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
	In this direction, one striking result was the proof that $\dnds$ is in fact predicted to be independent of $\Ne$ under very general cirumstances, namely [je propose d'être plus exhaustif sur l'ensemble des conditions des résultats de Cherry], whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable. Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}. The thermodynamic stability is itself computed using a 3D structural model of the protein. These simulation experiments have led to the observation that $\dnds$ is essentially independent of $\Ne$. The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.

	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype. For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur. More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability the protein.

	[Ici, il faut élaborer beaucoup plus en détail la question de l'articulation avec niveaux d'expression. Voici ce que je propose.]
	In addition, if empirical evidence for a negative correlation of $\dnds$ with $N_e$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\dnds$ with expression level or protein abundance. Indeed, expression level is one of the best predictors of $\dnds$, with highly expressed proteins typically having lower $\dnds$ values. Of note, the slope of the response of $\dnds$ to changes in expression level is relatively weak (ref Drummond), although clearly significant. Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\dnds$ and expression level are such that the response of $\dnds$ to changes in expression level or in $N_e$ should be very similar. Conversely, under strict equimutability, most of these models would predict that $\dnds$ should be essentially independent, not just of $N_e$, but also of expression level, thus at odds with empirical observations.

	[La aussi, il faut ressaisir l'ensemble de la question. Il y a plusieurs enjeux: il faut se positionner de façon claire, mais diplomatique, par rapport aux résultats et interprétations déjà proposés, par Cherry, Goldstein et d'autres: concéder que la relation omega Ne est faible, mais pas inexistante, et devant être quantifiée. D'autre part, il faut commencer à dessiner une perspective qui justifie le travail que tu vas présenter, et bien souligner comment ce travail théorique pourrait s'articuler avec les questionnements empiriques.]
	Overall, there is no doubt that the impact of changes in $N_e$ (or in expression level) on $\dnds$ is at best weak, making empirical estimation more difficult. Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful. [Ultimately, it could be confronted with more refined empirical estimates of the relation between $\dnds$ and both $N_e$ and expression level - à développer ici: rcomment la pente dépend de paramètres structuraux, et du coup, est potentiellement informative au sujet de ce qui détermine ultimement le paysage sélectif].

	Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium. However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection in the long term \cite{Jones2016}.
	Studying only equilibrium properties can thus be misleading. For that reason, the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized [là aussi, justifier l'articulation entre compréhension théorique et impact potentiel sur l'analyse empirique].

	%Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
	[Et maintenant seulement, tu peux décliner le contenu du manuscrit.] The goal of this study is to characterize how the response of $\dnds$ to changes in $\Ne$ quantitatively depends on the intrinsic [structural?] molecular parameters of the model. To this end, we develop a general mathematical approach for deriving the $\dnds$ relationship to $\Ne$ in the context of a given genotype-phenotype-fitness map. [Based on these results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data -- Me paraît être dit de façon un peu trop brutale, et trop naive aussi: aucun doute que la contrainte conformationnelle est une composante essentielle de la sélection purificatrice sur les protéines. La question est plutôt de savoir dans quelle mesure les éventuelles modulations du dnds en fonction de Ne dépendent de cette composante-là de la fitness, ou alors d'autres enjeux biophysiques, ou autres. Pourrais dire quelque chose d'un peu plus indirect, par exemple: we briefly discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\dnds$ and $N_e$. Quoi qu'il en soit, la fin de ce paragraphe est encore incomplète et pourra être revue une fois que la discussion aura été complétée. ]
	% The assumptions are strong, we want to relax equimutability which is unlikely

	\section*{Results}

	\subsection*{Models of evolution}

	[ Il faut présenter le cadre général: deux niveaux de modèles: modèle abstrait, modèle thermo, et justifier leur articulation].
	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e. such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$. As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability. This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations. 
	%Throughout our derivation, we present the results both for the general case and for the more specific example.

	In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.

	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues. Instead, each site contributes independently to $\deltaG$. 
	With such approximation, at a given site of sequence each amino-acid is either stabilizing or destabilizing the protein, and an unstable amino-acid contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ compare to a stable amino-acid. [dire plus explicitement: en chaque position, un acide aminé optimal, tous les autres considérés comme également destabilisateurs].
	If all amino-acids of the sequence are stable [je dirais plutôt: an amino-acid is stabilizing / destabilizing at a given position, plutôt que stable / unstable: the protein is stable / unstable, not the amino acid], $\deltaG$ is at its minimum, equal to $ \alpha < 0$. 
	In this model, the phenotype of a given genotype (i.e. sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq \x \leq 1$. Thus, $\deltaG$ is a linear function of $\x$:
	\begin{align}
	\deltaG (\x) = \alpha + \Nsite \gamma \x,
	\end{align}
	where $\Nsite$ is the number of sites in the sequence. \\

	[Some words about the phenotype - fitness map]. The fitness is assumed to be equal to the proportion of protein in folded conformation. This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
	This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
	\begin{align}
	f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
	\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
	\end{align}
	where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.

	\subsection*{Elasticity of $\dnds$ to changes in $N_e$. Analytical approximation}

	[Vu que c'est le résultat essentiel, je propose de mettre la notion d'élasticité beaucoup plus clairement en avant.]
	In this section we present an analytical approximate solution for the change in [response of equilibrium] $\dnds$ at equilibrium after a change in $\Ne$. By analogy with thermodynamics, we call this reponse the \emph{elasticity} of $\dnds$ to changes in $N_e$, and denote it as $\chi$:
	\begin{align}
	\chi = \frac{ \der \dnds}{\der \ln (\Ne)}
	\end{align}
	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein [in the specific case of the biophysical model?].
	All development are available in Supplementary Materials in the most general case.

	[Paragraphe à revoir en fonction du changement de contexte opéré par le fait que le modèle est maintenant décrit dans la section précédente.]
	The first step is to define the relationship from genotype to fitness, through an intermediary phenotype
	The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
	Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation.
